---
title: 'Lab 1: Linear models for quantitative genetics'
subtitle: "BMI 206"
author: "Valeria Sanchez Estrada"
date: "10/18/2024"
output: html_document
---

### PART1: Analyzing provided genotype and phenotype data.

**Prepare the data.** Read in the genotype and phenotype matrices.

```{r}
genos = as.matrix(read.table("./genos.txt"))
phenos = as.matrix(read.table("./phenos.txt"))
```

<br>

Make a histogram of the phenotypes. Do they look normally distributed?

Yes! Slightly skewed left.

```{r}
hist(phenos)
```

<br>

How are the genotypes encoded?

0, 1, and 2. For ex: AA, AG, GG

```{r}
table(genos)
```

<br>

How many individuals are there in the dataset and how many SNPs? (Save them in `N` and `M`, respectively.)

```{r}
dim(genos)
dim(phenos)
N = dim(genos)[1]
M = dim(genos)[2]
```

<br>

**Compute the *minor* allele frequency for every SNP. Check MAFs are \<0.5.**

```{r, eval=FALSE}
MAFs = array(0,M)
for(i in 1:M) {
      loc <- genos[,i]
      allele_1 <- sum(loc==0)*2 + sum(loc==1)
      allele_2 <- sum(loc==2)*2 + sum(loc==1)
      MAFs[i] = min(allele_1, allele_2) / (2 * N)
}
MAFs[1:10]
max(MAFs)
```

<br>

**Run a GWAS under an additive model and save the p-values, z-scores, and effect sizes.**

```{r}
pvalues = array(0,M)
zscores = array(0,M)
betas = array(0,M)
for(i in 1:M) {
	g = genos[,i]
	res = summary(lm(phenos~g))
	zscores[i] = res$coefficients[2, 'Estimate']/res$coefficients[2, 'Std. Error']
	pvalues[i] = res$coefficients[2, 'Pr(>|t|)']
	betas[i] = res$coefficients[2, 'Estimate']
}
```

<br>

Summarize the effect sizes.

```{r}
summary(betas)
hist(betas)
```

<br>

Are there any significantly associated SNPs? If so, which SNPs are they?

```{r}
pvalues=p.adjust(pvalues, method='bonferroni')
assoc = which(pvalues<=0.05)
assoc
```

<br>

How big are their effect sizes? How significant are they?

```{r, eval=FALSE}
betas[assoc]
zscores[assoc]
pvalues[assoc]
```

<br>

Draw a QQ plot for log10(p) values.

```{r}
obsLogPvs = sort(-log10(pvalues))
expLogPvs = sort(-log10(seq(1/M,1,1/M)))
plot(expLogPvs,obsLogPvs,main='QQ plot')
abline( a=0, b=1 )
#label the significant SNPs red 
points(expLogPvs[(M-length(assoc)):M],obsLogPvs[(M-length(assoc)):M],col="red")
```

<br>

Is there inflation? Use the chi-square statistics to check.

No, value is very close to 1.

```{r, eval=FALSE}
chis = zscores^2
lambdaGC = median(chis)/0.454 # why .454? median of a chi-squared distribution with one degree of freedom
lambdaGC # want value that is = 1
```

<br>

Plot the phenotype predictions for the most significant SNP.

```{r}
topSNP = genos[,order(pvalues)[1]]
plot(topSNP,phenos)
abline(lm(phenos~topSNP)$coeff,col="red")
```

<br>

**Build a linear predictor of the phenotype using the associated SNPs.**

```{r}
ypred = array(0,N)
for(i in 1:N) {
      ypred[i] = genos[i,assoc] %*% betas[assoc]
}
plot(ypred,phenos)
```

<br>

What is the correlation between the predicted phenotype and the true phenotype?

```{r, eval=FALSE}
cor(ypred,phenos)
```

<br>

**BONUS: Test each of the associated SNPs for non-linearity.**

```{r}
hp = array(0,length(assoc))
for (i in 1:length(assoc)) {
  g = genos[,assoc[i]]
  h = g
  h[h==2]=0
  #Hint: can use anova(lm(?),lm(?)) or summary(lm(?))
  hp[i] <- summary(lm(phenos~h))$coefficients[2, "Pr(>|t|)"] #skip multiple test correction for now
}
```

<br>

**BONUS: Visualize a linear SNP and a non-linear SNP.**

```{r}
par( mfrow=c(1,2) )
plot(genos[,assoc[which(hp<=0.05)[1]]], phenos, main='Non-Linear SNP' )
points( c(0,1,2), tapply(phenos, genos[,assoc[which(hp<=0.05)[1]]], mean ), col=2, pch=16, cex=3 )
lines( c(0,1,2), tapply(phenos, genos[,assoc[which(hp<=0.05)[1]]], mean ), col=2, lwd=2  )
plot(genos[,assoc[which(hp>0.05)[1]]], phenos, main='Linear SNP' )
points( c(0,1,2), tapply(phenos, genos[,assoc[which(hp>0.05)[1]]], mean ), col=2, pch=16, cex=3 )
lines( c(0,1,2), tapply(phenos, genos[,assoc[which(hp>0.05)[1]]], mean ), col=2, lwd=2  )
```

<br>

**Repeat the GWAS to test for recessive rather than additive genetic effects.**

```{r}
genos2 = genos
genos2[genos<2]=1
pvalues2 = array(0,M)
zscores2 = array(0,M)
betas2 = array(0,M)
for(i in 1:M) {
  g = genos2[,i]
  res = summary(lm(phenos~g))
  zscores2[i] = res$coefficients[2, 'Estimate']/res$coefficients[2,'Std. Error']
  pvalues2[i] = res$coefficients[2, 'Pr(>|t|)']
  betas2[i] = res$coefficients[2, 'Estimate']
}
```

<br>

**Are the same SNPs significant or not?**

No, SNP 10 is now not significant.

```{r, eval=FALSE}
pvalues2 <- p.adjust(pvalues2, method='bonferroni')
assoc2 = which(pvalues2<0.05)
assoc2
```

<br>

**How did the effect sizes change?**

It seems like a lot of the SNPs that previously had small effect sizes now either have much stronger effects (both positive and negative) on the phenotype.

```{r}
plot(betas,betas2)
```

<br>

### PART2: Simulating genotypes with LD.

**Establish some important simulation parameters.**

```{r}
N = 1000 #number of individuals
M = 30   #number of non-causal SNPs
gs = matrix(0,nrow=N,ncol=M)
```

<br>

**Simulate a GWAS data set.** First, simulate the causal variant.

```{r}
set.seed(42) #set random seed so we all get the same numbers
MAF = 0.5
gC = rbinom(N,1,MAF) #causal variant
```

<br>

Then, simulate the phenotypes given the causal variant.

```{r}
beta = 0.3 #association of causal variant
pheno = gC*beta + rnorm(N) 
```

<br>

Generate 10 SNPS in tight LD with the causal SNP.

```{r}
rho = 0.9
for(i in 1:10) {
  idx = rbinom(N,1,rho)
  gs[,i]=gC*idx+rbinom(N,1,MAF)*(1-idx)
  # test they have the right LD empirically
  cat( 'Observed LD = ', cor( gs[,i], gC ), '\n' )
  # Bonus: prove they have the right LD theoretically
}
```

<br>

Do the same for 10 moderate LD partners (rho=0.6).

```{r}
rho2 = 0.6
for(i in 11:20) {
  idx = rbinom(N,1,rho2)
  gs[,i]=gC*idx+rbinom(N,1,MAF)*(1-idx)
  # test they have the right LD empirically
  cat( 'Observed LD = ', cor( gs[,i], gC ), '\n' )
  # Bonus: prove they have the right LD theoretically
}
```

<br>

Do the same for 10 independent SNPs (rho=0).

```{r}
rho3 = 0
for(i in 21:30) {
  idx = rbinom(N,1,rho3)
  gs[,i]=gC*idx+rbinom(N,1,MAF)*(1-idx)
  # test they have the right LD empirically
  cat( 'Observed LD = ', cor( gs[,i], gC ), '\n' )
  # Bonus: prove they have the right LD theoretically
}
```

**Run GWAS on the causal variant. Then run GWAS on the other variants. Keep track of the zscores only.**

```{r}
zsC = summary(lm(pheno~gC))$coef[2,3]
zs = sapply( 1:M, function(i) summary(lm(pheno~gs[,i]))$coef[2,3] )
```

<br>

Visualize the relationship between the mean z-scores at the tag SNPs and the z-score at the causal SNP.

```{r}
par( mfrow=c(2,2) )
breaks = hist(c(0,zsC,zs),plot=F)$breaks
hist(zs[1:10],breaks=breaks, col=1, main='LD partners')
abline(v = zsC)
hist(zs[11:20],breaks=breaks, col=2, main='Low-LD partner SNPs')
abline(v = zsC)
hist(zs[21:30],breaks=breaks, col=3, main='Independent SNPs')
abline(v = zsC)
```

<br>

**BONUS: Perform LD score regression. First, calculate the LD scores. There should be M+1 of them.**

```{r, eval=FALSE}
genotype_matrix <- cbind(gC, gs)
ldscores <- numeric(ncol(genotype_matrix))
for (i in 1:ncol(genotype_matrix)) {
  for (j in 1:ncol(genotype_matrix)) {
    if (i != j) {
      r2 <- (cor(genotype_matrix[,i], genotype_matrix[,j]))^2
      ldscores[i] <- ldscores[i]+r2
    }
  }
}
ldscores
```

<br>

**BONUS: Visualize LD score regression.**

```{r,eval=FALSE}
chis = c(zs, zsC)^2
plot(ldscores, chis, ylab=expression(chi^2) )
#test for inflation
lambdaGC = median(chis)/0.454
lambdaGC
```

<br>

**BONUS: Estimate heritability.**

```{r,eval=FALSE}
summary(lm(pheno ~ gC))$coef[2,1] * M / N
```

<br>

**BONUS: What is the true heritability?**

```{r, eval=FALSE}
var(gC*beta) / var(pheno)
```
