---
title: 'Lab 1: Linear models for quantitative genetics'
author: "Zoe Chen"
date: "Oct 18 2024"
output:
  html_document: default
  pdf_document: default
subtitle: BMI 206
---
<br>
<br>
### PART1: Analyzing provided genotype and phenotype data.

__Prepare the data.__
Read in the genotype and phenotype matrices. 
```{r}
genos = as.matrix(read.table("./genos.txt"))
phenos = as.matrix(read.table("./phenos.txt"))
```
<br>

Make a histogram of the phenotypes. Do they look normally distributed?

A:Yes, they look normally distributed. 
```{r}
hist(phenos)
```
<br>

How are the genotypes encoded?

A:They encoded as 0, 1, and 2. If we are using "A" and "G" here, 0 is "AA", 1 is "AG" and 2 is "GG".
```{r}
table(genos)
```
<br>

How many individuals are there in the dataset and how many SNPs? (Save them in `N` and `M`, respectively.)

A:There are 1500 individuals and 10000 SNPs. 
```{r}
dim(genos)
dim(phenos)
N = 1500
M = 10000
```

<br>

__Compute the *minor* allele frequency for every SNP. Check MAFs are <0.5.__
```{r}
MAFs = array(0,M)
for(i in 1:M) {
  #Minor Allele Frequency: compute the MAF by taking the minimum of the two allele frequencies because the MAF must always be the less common allele's frequency.
  # Calculate allele frequencie
  #A_freq <- (2*nrow(genos[genos[, i] == 0, ]) + nrow(genos[genos[, i] == 1, ])) / (2*nrow(genos))
  A_freq <- ((2*table(genos[, i])[1]) + table(genos[, i])[2])/3000
  G_freq <- 1 - A_freq
  
  # Minor Allele Frequency: select the minimum of the allele frequencies
  MAFs[i] = min(A_freq, G_freq)
}
MAFs[1:10]
max(MAFs)
```
<br>

__Run a GWAS under an additive model and save the p-values, z-scores, and effect sizes.__
```{r}
pvalues = array(0,M)
zscores = array(0,M)
betas = array(0,M)
for(i in 1:M) {
	g = genos[,i]
	res = summary(lm(phenos~g))
	betas[i] = res$coefficients[2, "Estimate"]
	zscores[i] = res$coefficients[2, "t value"]
	pvalues[i] = res$coefficients[2, "Pr(>|t|)"]

}

pvalues[1:10]
zscores[1:10]
betas[1:10]

```
<br>

Summarize the effect sizes.
```{r}
summary(betas)
hist(betas)
```
<br>

Are there any significantly associated SNPs? If so, which SNPs are they?

A: Yes, when I use corrected p-val, there are 10 significant SNPs. They are listed as below. 
```{r}
assoc = which(pvalues<(0.05/M))
assoc
```
<br>

How big are their effect sizes? How significant are they? 

A:The effect sizes range from -3.4007282 to 2.2455246. They are significant range from extreme p-val 3.488001e-86 to not that extreme 6.308608e-07, but all passed the threshold. 
```{r}
betas[assoc]
zscores[assoc]
pvalues[assoc]
```
<br>

Draw a QQ plot for log10(p) values.
```{r}
obsLogPvs = sort(-log10(pvalues))
expLogPvs = sort(-log10(seq(1/M,1,1/M)))
plot(expLogPvs,obsLogPvs,main='QQ plot')
abline( a=0, b=1 )
#label the significant SNPs red 
points(expLogPvs[(M-length(assoc)):M],obsLogPvs[(M-length(assoc)):M],col="red")
```
<br>

Is there inflation? Use the chi-square statistics to check.

A:0.454 = qchisq(0.5, df=1), that's the the expected median chi-square statistic (under the null hypothesis). Lambda GC is very close to 1 indicates no inflation in this case. 
```{r}
chis = zscores ^ 2
lambdaGC = median(chis)/0.454 # why .454?
lambdaGC
```
<br>

Plot the phenotype predictions for the most significant SNP.
```{r}
topSNP = genos[,order(pvalues)[1]]
plot(topSNP,phenos)
abline(lm(phenos~topSNP)$coeff,col="red")
```
<br>

__Build a linear predictor of the phenotype using the associated SNPs.__
```{r}
ypred = array(0,N)
for(i in 1:N) {
      ypred[i] = genos[i,assoc] %*% betas[assoc]
}
plot(ypred,phenos)
```
<br>

What is the correlation between the predicted phenotype and the true phenotype?

A:It's 0.958, they are highly correlated.
```{r}
cor(ypred,phenos)
```
<br>

__BONUS: Test each of the associated SNPs for non-linearity.__
```{r}
hp = array(0,length(assoc))
for (i in 1:length(assoc)) {
  g = genos[,assoc[i]]
  h = g
  h[h==2]=0
  #Hint: can use anova(lm(?),lm(?)) or summary(lm(?))
  hp[i] <- anova( lm(phenos~g), lm(phenos~g * h) )$Pr[2] #skip multiple test correction for now
}
hp 
```
<br>

BONUS: Visualize a linear SNP and a non-linear SNP.

A:Based on the anova results above, the 5th SNPs has better linear fit and the 1st SNPs is a more non-linear SNP
```{r}
par( mfrow=c(1,2) )
plot( genos[,5], phenos )
points( c(0,1,2), tapply( phenos,  genos[,5], mean ), col=2, pch=16, cex=3 )
lines( c(0,1,2), tapply( phenos,  genos[,5], mean ), col=2, lwd=2  )
plot( genos[, 1], phenos )
points( c(0,1,2), tapply( phenos, genos[, 1], mean ), col=2, pch=16, cex=3 )
lines( c(0,1,2), tapply( phenos, genos[, 1], mean ), col=2, lwd=2  )
```
<br>

__Repeat the GWAS to test for recessive rather than additive genetic effects.__
```{r}
genos2 = genos
genos2[genos<2]=1
pvalues2 = array(0,M)
zscores2 = array(0,M)
betas2 = array(0,M)
for(i in 1:M) {
  g = genos2[,i]
  res = summary(lm(phenos~g))
	betas2[i] = res$coefficients[2, "Estimate"]
	zscores2[i] = res$coefficients[2, "t value"]
	pvalues2[i] = res$coefficients[2, "Pr(>|t|)"]
}

betas2[1:10]
zscores2[1:10]
pvalues2[1:10]
```
<br>

__Are the same SNPs significant or not?__

A:Yes, 1-9 are still significant, 10 is not. 
```{r}
assoc2 = which(pvalues2<(0.05/10000))
assoc2
```
<br>

__How did the effect sizes change?__

A:The effect sizes for recessive deviated from the original additive effect. 
```{r}
hist(betas2)
plot(betas,betas2)
```
<br>

### PART2: Simulating genotypes with LD.

__Establish some important simulation parameters.__
```{r}
N = 1000 #number of individuals
M = 30   #number of non-causal SNPs
gs = matrix(0,nrow=N,ncol=M)
```
<br>

__Simulate a GWAS data set.__

First, simulate the causal variant.
```{r}
set.seed(42) #set random seed so we all get the same numbers
MAF = 0.5
gC = rbinom(N,1,MAF) #causal variant
```
<br>

Then, simulate the phenotypes given the causal variant.
```{r}
beta = 0.3 #association of causal variant
pheno = gC*beta + rnorm(N) 
```
<br>

Generate 10 SNPS in tight LD with the causal SNP.

A：For bonus to theoretically prove they have the right LD, we should prove r^2 is close to the expected rho^2
```{r}
set.seed(42)
rho = 0.9
for(i in 1:10) {
  idx = rbinom(N,1,rho)
  gs[,i]=gC*idx+rbinom(N,1,MAF)*(1-idx)
  # test they have the right LD empirically
  cat( 'Observed LD = ', cor( gs[,i], gC ), '\n' )
  
  
  # Bonus: prove they have the right LD theoretically
  prob_gC = mean(gC)  # Causal variant allele frequency
  prob_snp = mean(gs[,i])  # Linked SNP allele frequency
  # Calculate r^2
  prob_gC1_snp1 = mean((gC == 1) & (gs[,i] == 1))  # p(A1B1)
  D = prob_gC1_snp1 - (prob_gC * prob_snp)
  r2 = D^2 / (prob_gC * (1 - prob_gC) * prob_snp * (1 - prob_snp))
  # Check if r^2 is close to the expected rho^2
  print(abs(r2 - rho^2) < 0.1)
}
```
<br>

Do the same for 10 moderate LD partners (rho=0.6).
```{r}
set.seed(42)
rho = 0.6
for(i in 11:20) {
  idx = rbinom(N,1,rho)
  gs[,i]=gC*idx+rbinom(N,1,MAF)*(1-idx)
  # test they have the right LD empirically
  cat( 'Observed LD = ', cor( gs[,i], gC ), '\n' )
  
  
  
  # Bonus: prove they have the right LD theoretically
  prob_gC = mean(gC)  # Causal variant allele frequency
  prob_snp = mean(gs[,i])  # Linked SNP allele frequency
  # Calculate r^2
  prob_gC1_snp1 = mean((gC == 1) & (gs[,i] == 1))  # p(A1B1)
  D = prob_gC1_snp1 - (prob_gC * prob_snp)
  r2 = D^2 / (prob_gC * (1 - prob_gC) * prob_snp * (1 - prob_snp))
  # Check if r^2 is close to the expected rho^2
  print(abs(r2 - rho^2) < 0.1)
}
```
<br>

Do the same for 10 independent SNPs (rho=0).
```{r}
set.seed(42)
rho = 0
for(i in 21:30) {
  idx = rbinom(N,1,rho)
  gs[,i]=gC*idx+rbinom(N,1,MAF)*(1-idx)
  # test they have the right LD empirically
  cat( 'Observed LD = ', cor( gs[,i], gC ), '\n' )
  
  
  
  # Bonus: prove they have the right LD theoretically
  prob_gC = mean(gC)  # Causal variant allele frequency
  prob_snp = mean(gs[,i])  # Linked SNP allele frequency
  # Calculate r^2
  prob_gC1_snp1 = mean((gC == 1) & (gs[,i] == 1))  # p(A1B1)
  D = prob_gC1_snp1 - (prob_gC * prob_snp)
  r2 = D^2 / (prob_gC * (1 - prob_gC) * prob_snp * (1 - prob_snp))
  # Check if r^2 is close to the expected rho^2
  print(abs(r2 - rho^2) < 0.1)
}
```

__Run GWAS on the causal variant. Then run GWAS on the other variants. Keep track of the zscores only.__
```{r}
zsC = summary(lm(pheno~gC))$coef[2,3]
zs = sapply( 1:M, function(i) summary(lm(pheno~gs[,i]))$coef[2,3] )
```
<br>

Visualize the relationship between the mean z-scores at the tag SNPs and the z-score at the causal SNP.
```{r}
par( mfrow=c(2,2) )
breaks = hist(c(0,zsC,zs),plot=F)$breaks
hist(zs[1:10],breaks=breaks, col=1, main='LD partners')
abline(v=zsC)
hist(zs[11:20],breaks=breaks, col=2, main='Low-LD partner SNPs')
abline(v=zsC)
hist(zs[21:30],breaks=breaks, col=3, main='Independent SNPs')
abline(v=zsC)
```
<br>

__BONUS: Perform LD score regression. First, calculate the LD scores. There should be M+1 of them.__

A: by formula, the LD score = sum of the the squared correlations (also, the linkage disequilibrium) values between that SNP and all other SNPs.
```{r}
all_snps = cbind(gC, gs)

ldscores = numeric(M+1) 

# Calculate LD scores for all SNPs
for (i in 1:ncol(all_snps)) {
  # Calculate r
  r = cor(all_snps[, i], all_snps) 
  ldscore = sum(r^2)
  ldscores[i] = ldscore 
}

ldscores
```
<br>

BONUS: Visualize LD score regression.
```{r}
chis = c(zsC^2, zs^2)
plot( ldscores, chis, ylab=expression(chi^2) )
#test for inflation
lambdaGC = median(chis)/0.454
lambdaGC
```
<br>

BONUS: Estimate heritability.
```{r}
summary(lm(chis ~ ldscores))$coef[2,1] * M/N
```
<br>

BONUS: What is the true heritability?

A:heritability is calculated as the ratio of genetic variance to total phenotypic variance
```{r}
var(gC * beta) / var(pheno)
```









